---
title: "Dance Entropy Analysis"
author: "Peter Gates"
date: "10/30/2019"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate) # date time formatting
library(ggplot2)
library(ggthemes)
df <- read_csv("C:/Users/albei/Nextcloud/Documents/PhD/Ridgel Lab/Dance Poster/Entropy Analysis/Data/SamEn_joints.csv")
```

## Dance Study Entropy Analysis

Entropy has been calculated using MatLab code for 176 biomechanical variables measured through the Noraxon system during dance sessions in participants with Parkinson's Disease (PD) and control older adults (OA). This document outlines the statistical results. 

### Data Cleaning

Data was cleaned using the following criteria:

* Date was extracted from filename and added to new Date column
* Participant ID extracted from filename and added to new Participant column
* Dance Type was extracted from filename and added to new Dance Type column
* Rumba, Swing, Electric Slide data were removed
* New Group column was created where data was organized into "OA" or "PD"
* All values equal to "0" were replaced with "NA". 0 indicates a failure of the biomechanical sensors and thus are excluded from analysis
* One observation of Tango was removed as there were NA data for any of the variables.

The cleaned data were saved as df_clean.

df_clean was further separated into PD or OA data, for ease of analysis. These are shown below.


```{r clean, include=FALSE, echo=TRUE}
df_clean_path = "C:/Users/albei/Nextcloud/Documents/PhD/Ridgel Lab/Dance Poster/Entropy Analysis/Data/clean_samen.csv"

df_clean <- df %>%
  # Extract appropriate entropy
  select(-useless_col) %>%
  # Create Date column
  mutate(Date = str_extract(string = file_name, pattern = "^20..............")) %>%
  mutate(Date = ymd_hm(Date)) %>%
  # Create Participant column
  mutate(Participant = str_extract(file_name, "pddance[0-9][0-9][0-9]|pddancecon[0-9][0-9][0-9]")) %>%
  mutate(Participant = factor(Participant)) %>%
  # Create Dance_Type column
  mutate(file_name = str_remove(file_name,".*pddance[0-9][0-9][0-9]_")) %>%
  mutate(file_name = str_remove(file_name,".*pddancecon[0-9][0-9][0-9]_")) %>%
  mutate(Dance_Type = str_remove(file_name,".xlsx")) %>%  
  # Reorder, remove duplicates
  select(Date, Participant, Dance_Type, ends_with("(deg)")) %>% # EDIT ME
  # Clean Dance column
  mutate(Dance_Type = str_replace(Dance_Type, pattern = ".*Tango.*|.*tango.*", replacement = "Tango")) %>%
  mutate(Dance_Type = str_replace(Dance_Type, pattern = ".*Waltz.*|.*waltz.*", replacement = "Waltz")) %>%
  mutate(Dance_Type = str_replace(Dance_Type, pattern = ".*Waltz.*|.*waltz.*", replacement = "Waltz")) %>%
  mutate(Dance_Type = str_replace(Dance_Type, pattern = ".*line.*|.*Line.*", replacement = "Line Dance")) %>%
  mutate(Dance_Type = str_replace(Dance_Type, pattern = ".*foxtrot.*|.*Foxtrot.*", replacement = "Foxtrot")) %>%
  mutate(Dance_Type = str_replace(Dance_Type, pattern = ".*Prominent.*", replacement = "Tango")) %>%
  mutate(Dance_Type = str_replace(Dance_Type, pattern = ".*BoxStep.*", replacement = "Foxtrot")) %>%
  mutate(Dance_Type = str_replace(Dance_Type, pattern = ".*Electric slide.*|.*Electric Slide.*", replacement = "Electric_Slide")) %>%
  mutate(Dance_Type = str_replace(Dance_Type, pattern = ".*rumba.*|.*Rumba.*", replacement = "Rumba")) %>%
  mutate(Dance_Type = factor(Dance_Type)) %>%
  # Remove Rumba, Swing, Electric Slide
  filter(Dance_Type != "Rumba") %>%
  filter(Dance_Type != "Swing") %>%
  filter(Dance_Type != "Electric_Slide") %>%
  # Control v PD column
  mutate(Group = Participant) %>%
  mutate(Group = str_replace(Group, pattern = ".*pddancecon.*", replacement = "OA")) %>%
  mutate(Group = str_replace(Group, pattern = ".*pddance.*", replacement = "PD")) %>%
  mutate(Group = factor(Group))

df_clean[df_clean == 0] <- NA

write_csv(df_clean, path = df_clean_path)
```

```{r separate, include=FALSE, echo=TRUE}
df_pd_path = "C:/Users/albei/Nextcloud/Documents/PhD/Ridgel Lab/Dance Poster/Entropy Analysis/Data/pd_samen.csv"
df_oa_path = "C:/Users/albei/Nextcloud/Documents/PhD/Ridgel Lab/Dance Poster/Entropy Analysis/Data/oa_samen.csv"

df_pd <- df_clean %>%
  filter(Group == "PD") %>%
  drop_na()

write_csv(df_pd, path = df_pd_path)

df_oa <- df_clean %>%
  filter(Group == "OA") %>%
  drop_na()

write_csv(df_oa, path = df_oa_path)
```

``` {r heads}
df_pd
df_oa
```

### Data Statistics

A MANOVA was run to determine whether any significant differences exist between dance types in terms of the 10 variables. This turned out to be significant (p < 0.001).

```{r manova_dance, echo=FALSE}

dep_vars <- as.matrix(select(df_pd, ends_with("(deg)")))
df_pd.man <- manova(dep_vars ~ df_pd$Dance_Type)

summary(df_pd.man, test = "Pillai")
```

The dependent variables significantly different between dances include:

* SamEn_Elbow-LT-Flexion (p < 0.01)
* SamEn_Hip-RT-Abduction (p < 0.01)
* SamEn_Knee-LT-Flexion was slightly significant (p < 0.1)

The following only shows the significant results:

``` {r manova_summary, echo=FALSE}
# summary.aov(df_pd.man)

print("SamEn_Elbow-LT-Flexion")
df_pd_LT_elbow.aov <- aov(df_pd$`SamEn_Elbow-LT-Flexion (deg)` ~ df_pd$Dance_Type)
summary.aov(df_pd_LT_elbow.aov)
print("SamEn_Hip-RT-Abduction")
df_pd_RT_hip.aov <- aov(df_pd$`SamEn_Hip-RT-Abduction (deg)`~ df_pd$Dance_Type)
summary.aov(df_pd_RT_hip.aov)
print("SamEn_Knee-LT-Flexion")
df_pd_LT_knee.aov <- aov(df_pd$`SamEn_Knee-LT-Flexion (deg)` ~ df_pd$Dance_Type)
summary.aov(df_pd_LT_knee.aov)
```

TukeyHSD was run to determine which dances saw the significant difference in the appropriate SamEn. 

* SamEn in LT elbow flexion was significantly different between TANGO and FOXTROT (p = 0.001)
* SamEn in RT hip abduction was significantly different between TANGO and LINE DANCE (p = 0.005), LINE DANCE and FOXTROT (p = 0.046)
* SamEn in LT knee flexion was NOT significantly different between any dances (p > 0.1 for all comparisons)

```{r tukey, echo=FALSE}
print("SamEn_Elbow-LT-Flexion")
TukeyHSD(df_pd_LT_elbow.aov)

print("SamEn_Hip-RT-Abduction")
TukeyHSD(df_pd_RT_hip.aov)

print("SamEn_Knee-LT-Flexion")
TukeyHSD(df_pd_LT_knee.aov)
```

### Data Graphs

The bargraphs below demonstrate that for PD participants Foxtrot had the greatest mean SamEn of both left elbow flexion and left knee flexion, reenforcing our findings from the TukeyHSD run above.

PD participants also tended to show greater mean right hip abduction SamEn when compared to OA participants.


```{r graphs_pd_bar, echo=FALSE}
# KSU Theme: unify the design
ksu_theme = list(
  geom_col(position = position_dodge(0.9), color = "black"), # black gives the bars an outline. Dodge separates instead of default stack
  labs(x = "Dance Types", y = "Mean SamEn"),
  theme_classic(),  # gets rid of background lines
  theme(plot.title = element_text(hjust = 0.55)),
  scale_fill_manual(" ", values = c("OA" = "yellow", "PD" = "Blue")), # default colors at KSU
  ylim(0,0.15)
)

lt_elbow_plt <- ggplot(df_clean, aes(
  x = Dance_Type,
  y = df_clean$`SamEn_Elbow-LT-Flexion (deg)`,
  fill = Group )) + 
  ggtitle("Mean SamEn of LT Elbow Flexion") +
  ksu_theme

rt_hip_plt <- ggplot(df_clean, aes(
  x = Dance_Type,
  y = df_clean$`SamEn_Hip-RT-Abduction (deg)`,
  fill = Group )) + 
  ggtitle("Mean SamEn of RT Hip Abduction") +
  ksu_theme

lt_knee_plt <- ggplot(df_clean, aes(
  x = Dance_Type,
  y = df_clean$`SamEn_Knee-LT-Flexion (deg)`,
  fill = Group )) + 
  ggtitle("Mean SamEn of LT Knee Flexion") +
  ksu_theme

lt_elbow_plt
rt_hip_plt
lt_knee_plt
```

Plotted below are the SamEn of those variables that proved to be significantly different between dances in PD participants. Next to each plot is the OA equivalent. From here we can see that:

* PD participants had a much wider range in SamEn of left hip abduction than OA participants during Tango
* Both PD and OA participants had lower SamEn of left elbow flexion than right elbow flexion
* PD participants had a wider range in SamEn of left knee flexion than OA participants during Tango

```{r graphs_dot, echo=FALSE}
library(ggrepel)
pos <- position_jitter(width = 0.5, seed = 2) # standardizes randomness

elbow_plt <- ggplot(df_clean, aes(x = Dance_Type)) + 
  geom_point(position = pos,
             aes(y = df_clean$`SamEn_Elbow-RT-Flexion (deg)`,
                 color = "6532")
  ) +
  geom_point(position = pos,
             aes(y = df_clean$`SamEn_Elbow-LT-Flexion (deg)`,
                 color = "8965"),
  ) +
  labs(title = "SamEn of Elbow Flexion",
       x = "Dance Types", 
       y = "SamEn") +
  theme(plot.title = element_text(hjust = 0.55),
        legend.position = "right") +
  geom_vline(xintercept = 1.5) +
  geom_vline(xintercept = 2.5) +
  geom_vline(xintercept = 3.5) +
  ylim(0,0.2) +
  scale_color_manual(name = "Location",
                     labels = c("Right Elbow Flexion", "Left Elbow Flexion"),
                     values = c("red","black")) +
  facet_wrap(~ Group)

hip_plt <- ggplot(df_clean, aes(x = Dance_Type)) + 
  geom_point(position = pos,
             aes(y = df_clean$`SamEn_Hip-RT-Abduction (deg)`,
                 color = "6532")
  ) +
  geom_point(position = pos,
             aes(y = df_clean$`SamEn_Hip-LT-Abduction (deg)`,
                 color = "8965"),
  ) +
  labs(title = "SamEn of Abduction",
       x = "Dance Types", 
       y = "SamEn") +
  theme(plot.title = element_text(hjust = 0.55),
        legend.position = "right") +
  geom_vline(xintercept = 1.5) +
  geom_vline(xintercept = 2.5) +
  geom_vline(xintercept = 3.5) +
  ylim(0,0.2) +
  scale_color_manual(name = "Location",
                     labels = c("Right Hip Abduction", "Left Hip Abduction"),
                     values = c("red","black")) +
  facet_wrap(~ Group)

knee_plt <- ggplot(df_clean, aes(x = Dance_Type)) + 
  geom_point(position = pos,
             aes(y = df_clean$`SamEn_Knee-RT-Flexion (deg)`,
                 color = "6532")
  ) +
  geom_point(position = pos,
             aes(y = df_clean$`SamEn_Knee-LT-Flexion (deg)`,
                 color = "8965"),
  ) +
  labs(title = "SamEn of Knee Flexion",
       x = "Dance Types", 
       y = "SamEn") +
  theme(plot.title = element_text(hjust = 0.55),
        legend.position = "right") +
  geom_vline(xintercept = 1.5) +
  geom_vline(xintercept = 2.5) +
  geom_vline(xintercept = 3.5) +
  ylim(0,0.2) +
  scale_color_manual(name = "Location",
                     labels = c("Right Knee Flexion", "Left Knee Flexion"),
                     values = c("red","black")) +
  facet_wrap(~ Group)

hip_plt
elbow_plt
knee_plt
```